---
title: "AbiR + Olaparib treatment day 5 and 9: RNAseq summary"
author: "Marion Hardy"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true 
    theme: spacelab 
    highlight: monochrome
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = TRUE, echo = FALSE, warning = F, cache.lazy = F)
knitr::opts_chunk$set(fig.width=10, fig.height=15) 

library(tidyverse)
library(DESeq2)
library(cowplot)
library(ggrepel)
library(biomaRt)
library(RColorBrewer)
library(pheatmap)
library(xlsx)
library(msigdbr)
library(clusterProfiler)
library(kableExtra)

```

```{r Data loading, include=FALSE}

dds = readRDS("./data_output/AbiR_DESeq.Rds")

```

# Introduction

Abiraterone-resistant cancer cells (AbiR) were treated with:

-   DMSO and harvested on day 3 = DMSO Day 3
-   Olaparib 5 uM? and harvested on day 5 = Ola_5 Day 5
-   Olaparib 5 uM? and harvested on day 9 = Ola_5 Day 9

Was there a drug holiday in there? Need more context for this report.
Bulk RNAseq, all samples collected on the same day.

# Principal Component Analysis

```{r, fig.height=7, fig.width=5}

# Checking PCA

rld = vst(dds)

plotPCA(rld,intgroup="Condition")+
  theme_bw() +
  labs(title = 'PCA per condition')

```

Really good variance coverage by the first 2 components.
No batch effect observed. Most of the variance is explained by DMSO vs Ola.

# Euclidian Distance matrix

```{r, fig.height=5, fig.width=6}

# Checking sample similarity

sampleDists = dist(t(assay(rld)))
sampleDistMatrix = as.matrix(sampleDists)
rownames(sampleDistMatrix) = paste(rld$Condition, sep="-")
colnames(sampleDistMatrix) = paste(rld$Condition, sep="-")
colors = colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)

```

# Comparaisons of interest
## Picking comparisons based on AK's wishes

```{r}
resultsNames(dds) %>% 
  knitr::kable()
```

Maybe
- Ola day 5 vs DMSO
- Ola day 9 vs DMSO

## Olaparib day 5 vs DMSO

```{r}

res = results(dds, name = "Condition_Ola_Day_5_vs_DMSO")
res_tbl = as_tibble(res, rownames="ensembl")

# Import annotation file

ensembl_to_geneName = readRDS("./data/hsapiens_annotations_230510.rds")

# filter(!duplicated(gene, drop = F))

# Add gene names and entrez id to your results

res_tbl = res_tbl %>%
  left_join(ensembl_to_geneName) %>%
  arrange(padj) %>% 
  distinct(ensembl, .keep_all = T)
  

write.csv(res_tbl,"./data_output/Ola_Day_5_vs_DMSO/res_tbl.csv", row.names = T)

# Save the significant DEG in an excel file

overexpr = res_tbl %>% filter(padj<=0.05 & log2FoldChange>=1) 
underexpr = res_tbl %>% filter(padj<=0.05 & log2FoldChange<=-1)
signif = full_join(overexpr, underexpr)

write.xlsx(overexpr, file="./data_output/Ola_Day_5_vs_DMSO/Sign_genes.xlsx", sheetName="UP")
write.xlsx(underexpr, file="./data_output/Ola_Day_5_vs_DMSO/Sign_genes.xlsx", sheetName="DOWN", append=TRUE)
write.xlsx(signif, file="./data_output/Ola_Day_5_vs_DMSO/Sign_genes.xlsx", sheetName="SIGNIF", append=TRUE)

```

### Volcano plots

```{r, fig.height=6, fig.width=14}
# Volcano plot

p1 =
res_tbl %>%
  filter(!is.na(padj)) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj),
             color = padj < 0.05 & abs(log2FoldChange) > 1, 
             label = ifelse(padj<0.05&log2FoldChange>=1|
                              padj<0.05&log2FoldChange<=-1,as.character(gene),''))) +
  scale_colour_manual(values = c("gray", "firebrick3")) +
  labs(col="Significantly expressed")+
  geom_point(size = 0.5) +
  geom_hline(yintercept = -log10(0.05)) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1)+
  xlim(-7,9)+
  geom_text_repel(max.overlaps = 5,
                  box.padding = 0,
                  segment.color = 'grey80',
                  fontface = "italic")+
  labs(title = "Ola_Day_5_vs_DMSO")+
  theme_bw()

print(p1)

ggsave("./figures/Ola_Day_5_vs_DMSO/volcanoplot.png", last_plot(), 
       dpi= 500, width = 14, height = 6)

```

### GSEA analysis (logFc)

```{r}

ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

```

```{r}

# MsigDb

hsa_GOBP_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C5",
  subcategory = "GO:BP")

hsa_GOMF_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C5",
  subcategory = "GO:MF")

hsa_reactome_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C2",
  subcategory = "CP:REACTOME") # for reactome collection

hsa_kegg_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C2",
  subcategory = "CP:KEGG") # for KEGG collection

hsa_wiki_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C2",
  subcategory = "CP:WIKIPATHWAYS") # for Wikipathways collection

set.seed(054057) # for reproducibility

```

#### GOBP

```{r, fig.height=9, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
              subtitle = "GSEA GO:BP Ola day 5 vs DMSO")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  xlim(-3.2,2.2)

ggsave(plot = last_plot(), "./figures/Ola_Day_5_vs_DMSO/GOBPgsea.svg", dpi = 300,
       height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, "./data_output/Ola_Day_5_vs_DMSO/GOBPgsea.xlsx")

```

```{r}
gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```


```{r, fig.height=9, fig.width=9}

cnetplot(gsea_results, foldChange=ordered_genes_fc, showCategory = 30,
         cex_label_category = .5, cex_label_gene = .5, layout = 'fr')+
  scale_colour_gradient2(name = "fold change", 
                         low = "blue", 
                         mid = "white", 
                         high = "red")

```

#### GOMF

```{r, fig.height=9, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOMF_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 30, font.size = 7)+ 
  labs(title = "GSEA",
              subtitle = "GSEA GO:MF Ola day 5 vs DMSO")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  xlim(-3.2,2.2)

ggsave(plot = last_plot(), "./figures/Ola_Day_5_vs_DMSO/GOMFgsea.svg", dpi = 300,
       height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, "./data_output/Ola_Day_5_vs_DMSO/GOMFgsea.xlsx")

```

```{r}
gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

```{r, fig.height=15, fig.width=15}

cnetplot(gsea_results, foldChange=ordered_genes_fc, showCategory = 30,
         cex_label_category = .5, cex_label_gene = .5, layout = 'fr')+
  scale_colour_gradient2(name = "fold change", 
                         low = "blue", 
                         mid = "white", 
                         high = "red")
```

#### Reactome

```{r, fig.height=8, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 30, font.size = 7)+ 
  labs(title = "GSEA",
              subtitle = "GSEA Reactome Ola day 5 vs DMSO")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  xlim(-3.2,2.2)

ggsave(plot = last_plot(), "./figures/Ola_Day_5_vs_DMSO/Reactomegsea.svg", dpi = 300, height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, "./data_output/Ola_Day_5_vs_DMSO/Reactomegsea.xlsx")

```

```{r}
gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

```{r, fig.height=15, fig.width=15}

cnetplot(gsea_results, foldChange=ordered_genes_fc, showCategory = 30,
         cex_label_category = .5, cex_label_gene = .5, layout = 'fr')+
  scale_colour_gradient2(name = "fold change", 
                         low = "blue", 
                         mid = "white", 
                         high = "red")
```

#### Wikipathways

```{r, fig.height=8, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_wiki_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 30, font.size = 7)+ 
  labs(title = "GSEA",
              subtitle = "GSEA Wikipathways Ola day 5 vs DMSO")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  xlim(-3.2,2.2)

ggsave(plot = last_plot(), "./figures/Ola_Day_5_vs_DMSO/wikigsea.svg", dpi = 300,
       height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, "./data_output/Ola_Day_5_vs_DMSO/wikigsea.xlsx")

```

```{r}
gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

```{r, fig.height=15, fig.width=15}

cnetplot(gsea_results, foldChange=ordered_genes_fc, showCategory = 30,
         cex_label_category = .5, cex_label_gene = .5, layout = 'fr')+
  scale_colour_gradient2(name = "fold change", 
                         low = "blue", 
                         mid = "white", 
                         high = "red")
```

## Olaparib day 9 vs day 5

This will give you the differences between the two. Otherwise they are way to similar for you to be able to manually go through the minute changes.

```{r}

dds$Condition = relevel(dds$Condition, "Ola_Day_5")
dds = DESeq(dds)
resultsNames(dds)

res = results(dds, name = "Condition_Ola_Day_9_vs_Ola_Day_5")
res_tbl = as_tibble(res, rownames="ensembl")

# Add gene names and entrez id to your results

res_tbl = res_tbl %>%
  left_join(ensembl_to_geneName) %>%
  arrange(padj) %>% 
  distinct(ensembl, .keep_all = T)
  

write.csv(res_tbl,"./data_output/Ola_Day_9_vs_Ola_Day_5/res_tbl.csv", row.names = T)

# Save the significant DEG in an excel file

overexpr = res_tbl %>% filter(padj<=0.05 & log2FoldChange>=1) 
underexpr = res_tbl %>% filter(padj<=0.05 & log2FoldChange<=-1)
signif = full_join(overexpr, underexpr)

write.xlsx(overexpr, file="./data_output/Ola_Day_9_vs_Ola_Day_5/Sign_genes.xlsx", sheetName="UP")
write.xlsx(underexpr, file="./data_output/Ola_Day_9_vs_Ola_Day_5/Sign_genes.xlsx", sheetName="DOWN", append=TRUE)
write.xlsx(signif, file="./data_output/Ola_Day_9_vs_Ola_Day_5/Sign_genes.xlsx", sheetName="SIGNIF", append=TRUE)

```

### Volcano plots

```{r, fig.height=6, fig.width=14}
# Volcano plot

p1 =
res_tbl %>%
  filter(!is.na(padj)) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj),
             color = padj < 0.05 & abs(log2FoldChange) > 1, 
             label = ifelse(padj<0.05&log2FoldChange>=1|
                              padj<0.05&log2FoldChange<=-1,as.character(gene),''))) +
  scale_colour_manual(values = c("gray", "firebrick3")) +
  labs(col="Significantly expressed")+
  geom_point(size = 0.5) +
  geom_hline(yintercept = -log10(0.05)) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1)+
  xlim(-7,9)+
  geom_text_repel(max.overlaps = 5,
                  box.padding = 0,
                  segment.color = 'grey80',
                  fontface = "italic")+
  labs(title = "Ola_Day_9_vs_Ola_Day_5")+
  theme_bw()

print(p1)

ggsave("./figures/Ola_Day_9_vs_Ola_Day_5/volcanoplot.png", last_plot(), 
       dpi= 500, width = 7, height = 6)

```

### GSEA analysis (logFc)

```{r}

ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

```

#### GOBP

```{r, fig.height=9, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
              subtitle = "GSEA GO:BP Ola day 9 vs day 5")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  xlim(-3.2,2.2)

ggsave(plot = last_plot(), "./figures/Ola_Day_9_vs_Ola_Day_5/GOBPgsea.svg", dpi = 300,
       height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, "./data_output/Ola_Day_9_vs_Ola_Day_5/GOBPgsea.xlsx")

```

```{r}
gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```


```{r, fig.height=9, fig.width=9}

cnetplot(gsea_results, foldChange=ordered_genes_fc, showCategory = 30,
         cex_label_category = .5, cex_label_gene = .5, layout = 'fr')+
  scale_colour_gradient2(name = "fold change", 
                         low = "blue", 
                         mid = "white", 
                         high = "red")

```

#### GOMF

```{r, fig.height=9, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOMF_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 30, font.size = 7)+ 
  labs(title = "GSEA",
              subtitle = "GSEA GO:MF Ola day 9 vs day 5")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  xlim(-3.2,2.2)

ggsave(plot = last_plot(), "./figures/Ola_Day_9_vs_Ola_Day_5/GOMFgsea.svg", dpi = 300,
       height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, "./data_output/Ola_Day_9_vs_Ola_Day_5/GOMFgsea.xlsx")

```

```{r}
gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

```{r, fig.height=15, fig.width=15}

cnetplot(gsea_results, foldChange=ordered_genes_fc, showCategory = 30,
         cex_label_category = .5, cex_label_gene = .5, layout = 'fr')+
  scale_colour_gradient2(name = "fold change", 
                         low = "blue", 
                         mid = "white", 
                         high = "red")
```

#### Reactome

```{r, fig.height=8, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
    gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 30, font.size = 7)+ 
  labs(title = "GSEA",
              subtitle = "GSEA Reactome Ola day 9 vs day 5")+
   theme(axis.line = element_line(color='black'),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())+
  xlim(-3.2,2.2)

ggsave(plot = last_plot(), "./figures/Ola_Day_9_vs_Ola_Day_5/Reactomegsea.svg", dpi = 300, height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, "./data_output/Ola_Day_9_vs_Ola_Day_5/Reactomegsea.xlsx")

```

```{r}
gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

```{r, fig.height=15, fig.width=15}

cnetplot(gsea_results, foldChange=ordered_genes_fc, showCategory = 30,
         cex_label_category = .5, cex_label_gene = .5, layout = 'fr')+
  scale_colour_gradient2(name = "fold change", 
                         low = "blue", 
                         mid = "white", 
                         high = "red")
```

```{r}
sessionInfo()
```
